Detection of slow slip events using wavelet analysis of GNSS recordings

### 2022-04-09 paper/answer_reviewers.tex

\documentclass[letterpaper, 12pt]{article}

\usepackage{caption}
\usepackage{float}
\usepackage[T1]{fontenc}
\usepackage[lmargin=1 in, rmargin=1 in, tmargin=1 in, bmargin=1 in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{times}
\usepackage{xcolor}

\begin{document}

\textbf{Reviewer 1}

\bigskip

\textit{In "Detection of slow slip events using wavelet analysis of GNSS recordings", the authors propose a technique to detect slow slip events using the Maximal Overlap Discrete Wavelet Transform (MODWT). While previous authors have used or advocated the use of wavelets for the detection of slow slip in geodetic time series, this manuscript provides a rubric for detection and provides a connection to slow slip duration. Applying this rubric to slow slip events in northern Cascadia, the authors provide a method for using the MODWT on both geodetic and tremor duration time series to demonstrate the general applicability of the technique. The authors advocate using this technique in situations where the connection between slow slip and tectonic tremor is less obvious.} \\

\textit{This manuscript provide a useful and much needed exploration and explanation of the relationship between wavelet levels and offset duration. In particular, advocating for the use of the MODWT over the standard discrete wavelet transform provides a nice way to avoid boundary effects at the edge of each wavelet octave and provides better temporal location for identification of wavelet peaks. The applicability to the case of Cascadia presented in the manuscript is clear and concise as well. My one big wish for this manuscript is a demonstration of its power in the situations described in the Introduction (lines 69-74), in particular, for those subduction zones where the connection between tremor and slow slip is not straightforward. To that end, I think the biggest test missing from this manuscript is a test in one of the problematic subduction zones, such as New Zealand or Alaska. Much of the demonstration of the power of this technique lies in selecting the threshold. For instance, how applicable are the thresholds determined from northern Cascadia data to other regions (say NZ or AK)? If those subduction zones were examined, how would one go about selecting a threshold? In short, is it possible to select a threshold in the absence of a catalog or does threshold selection depend on slow slip events already being know from other methods?}

\bigskip

We added an analysis of New Zealand GPS data.

\bigskip

\textit{While Figure 10 does show that nearly all the thresholds examined demonstrate a higher true positive rate than false positive, not all threshold value have equal identification power. Additionally, the actual value of the threshold is missing from the diagram (it would be a third dimension in the diagram). It is therefore difficult to ascertain whether the threshold levels explored provide complete coverage of reasonable values.}

\bigskip

We modified Figure 10 to better show the results corresponding to the different values of the threshold, by using a different color for each point depending on the values taken by the two thresholds.

\bigskip

\textit{Thus, while I feel that this manuscript highlights a potentially useful mathematical technique for identifying slow slip in geodetic time series, its broader applicability is not demonstrated. This manuscript would be improved through a demonstration of the proposed method, along with a comparison to an existing slow slip catalog, in a separate slow slip area.}

\bigskip

We added a comparison of slow slip events detected with this method and slow slip events detected by Todd and Schwartz (2016) in New Zealand.

\bigskip

\textit{In addition, I note the following items:} \\

\textit{(Line 172): I'm pretty sure the reference frame for the PANGA GPS time series is ITRF14. Orbits from JPL in the IGb08 reference frame end in early-mid 2018 and the time series for PGC5 shown in Figure 3 extend at least half-way through 2020. Therefore, the orbits used to process PGC5 can't be in IGb08.}

\bigskip

Thank you for pointing out this mistake. We corrected the sentence.

\bigskip

\textit{(Line 391) If the time series is categorized by the values -1, 0, and 1, wouldn't that be a trinary time series, not a binary time series?}

\bigskip

We replaced binary by trinary.

\bigskip

\textit{(Lines 458-460) These lines discuss a small displacement in the time series at 2007.5, but I don't see it. Should I be looking at Figure 9? If so, which latitude band should I look at and which time series are they composed of? That would be the place to look for causal offsets.}

\bigskip

\textit{(Figure 2) The caption text is not clear, in particular the phrase "for two locations of the missing values" is vague. The text (lines 290-295) does provide a better explanation of this figure however. Rewording the caption to provide a more self-contained explanation would improve the caption. Also, the values on the time axis for the right hand column are difficult to read as formatted.}

\bigskip

We modified the caption and the values on the time axis.

\bigskip

\textit{(Figure 5) The ordering of the data and the detail levels in this figure are opposite that of the Figures 1-3. For the sake of comparison, these should all be in the same order.}

\bigskip

We reordered the data and the detail levels.

\bigskip

\textit{(Figure 6) When producing the stacked time series, is there overlap/redundancy? Looking at the map in Figure 4 suggests that most triangles will be made with an overlapping number of stations. How will this affect the mapping of strange station behavior? In particular, could the displacement discussed on Lines 458-460 all be caused by just one misbehaving station that shows up in multiple bins?}

(Figure 7) Why do the obvious red-to-blue transitions in the tremor data seem to disappear from the detail level 7 figure, but are present in the detail level 8 (Figure 6) and detail level 6 (Figure 8) data? Can anything be interpreted from that pattern?

(Figure 9) When stacking levels 6, 7 and 8, won't the lower numbered levels dominate because of their (generally) larger amplitudes (as seen in Figure 3)? Is it valuable to normalize the levels before summing, or will this cause other issues?

(Figure 10) Is it possible to select a threshold in the absence of a catalog or does threshold selection depend on slow slip events already being know from other methods?

(Figure 11) What are the dates selected by this algorithm? How do they compare with the dates given in Michel et al., 2019? Most look spot on, but others, in particular the 2009 event look like they are quite distant in time.

There are some awkward sentences as well:

 (Line 190-191): "where n is some integer greater than or equal to" sounds better than "higher or equal".

(Line 199): The terminology "details" and "smooths" sounds awkward, is that the standard terminology in the wavelet literature?

(Line 287-288) The sentence "The straight line starts at and ends at." is incomplete.

\textbf{Reviewer 2}

The authors present a wavelet-based method to identify slow slip events within the GNSS position time series. Rather than looking at the geodetic record of slow slip as a sum of periodic basis functions, the wavelet method allows for the isolation of slow slip signals across different time scales at different times during the period of study. The approach is of interest to the community and I think that this manuscript will eventually be publishable.

That said, my major concern is that the method as presented will be very difficult to apply to another region of study, especially one without a systematically generated catalog of tectonic tremor or low-frequency earthquakes. Here the authors validate the detections (i.e. establish a detection threshold) within the geodetic data with detections made in the seismic data. Looking at figures 6-8, there are many wiggles within the time series, and it's not obvious to me how to interpret these signals without the well-established catalogs of A. Wech and the PNSN (tremor) and Michel et al. (2019) (slow slip); such detailed and long-duration catalogs are not available outside of Japan to validate the results of the presented method. Aggravating my concern is the premise that the authors themselves set up in the introduction of the manuscript, that a method is needed to detect slow slip when tremor is not available as an independent control; the presented method does not reach that
goal. I would also highlight that slow slip in northern Cascadia is some of the most well behaved slow slip across all subduction zones, with short-ish recurrence intervals and consistently clear geodetic signals captured by a relatively dense GNSS network, that will likely show up more reliably than elsewhere with this sort of approach.

The figures for the most part are clear, but there are some improvements to be made (see comments below). I also find that the introduction is lacking details and there are way too few articles cited for background reading for interested readers (cf. first paragraph with only two unique citations that discusses observations made across many subduction zones).

I think a revised version of this manuscript that has reframed the purpose of the approach will be an interesting contribution that is appropriate to be published in BSSA. I list other minor comments below:

general - There is no discussion of common mode signals that will surely stack constructively across GNSS stations. The removal of sinusoidal seasonal signals will remove some common mode signal, but there can still exist some non-tectonic signals that will stack constructively within the bandwidth of interest here. There is a reasonable chance that this will have little impact on the results, but it should at least be verified/discussed given the proposed processing.

general - I would avoid using the term ETS as it only adds confusion, as lines 69-74 highlight. I would suggest using "slow slip" to describe geodetic observations of aseismic slip and "tectonic tremor" for the accompanying seismic signals. The following paragraph further underlines this potential confusion: discussing the moment release of an "ETS event" is ambiguous. As described, it is not obvious to those unfamiliar with slow slip and tremor that the aseismic moment of slow slip dominates the seismic moment of tremor, because ETS conflates the geodetic observations with the seismic ones. I would simply use the term slow slip when referring to the dominant aseismic transient.

general - I'm not sure that the "true detection" terminology used here is quite appropriate. This implies that we have an absolute knowledge of what's happening on the plate interface, and that there can't be new detections made here without already be corroborated by other catalogs. Perhaps "robust detection" might be more appropriate? In any case, this underlines my above major comment that this method is difficult to interpret without other catalogs.

line 42 - "feature" is a strange word choice and does not tell the reader much at all

line 60 - "it is more abundant in some places" doesn't tell us much

line 64 - "for which" -> whose

line 67 - Obara (2002) did not discuss slow slip, only tectonic tremor. I would suggest something like Obara et al., GRL, 2004

line 126 - "levels" does not mean very much without context

lines 148 - "details" same comment as above; this general term doesn't mean anything without context

line 151-152 - why is there supposed to be a difference between the two tidal gauges and what does that mean? Not clear without the reader being forced to go and read the cited article.

line 156 - I assume longitudinal displacements is supposed to mean east-west motion?

line 287-288 - missing text here

figure 2 - I would say figure 2 could be moved to the appendix or supplementary material. It's not key to the message of the manuscript

figures 6-8 - I think this figure would be more easily interpretable as well as much effective if only one panel was used, with alternating geodetic and seismic data for a given location (red triangle from Figure 4)in two different colors (black and something else). The red/blue rectangles that indicate detections could be four similar colors for geodetic/seismic detections (blue/purple for negative peaks and red/orange for positive peaks for example). I'd also suggest that figures 7 and 8 could be made into supplementary figures as they show similar information to figure 6

figure 7 - I'm not sure I understand why the threshold selected here is less effective (there are less events identified) than for levels levels 6 and 8. What could explain this?

figure 10 - I think this figure would be a bit clearer if the 6/7/8 detail levels were in different colors

\end{document}

### 2022-04-09 paper/paper_bssa_annotated.tex
Work on revisions

Slow slip events are a new feature discovered in the last two decades in many subduction zones thanks to recordings of the displacement of Earth's surface by dense Global Navigation Satellite System (GNSS) networks ~\citep{VER_2010,SCH_2010,JIA_2012,WAL_2012}. As with ordinary earthquakes, slow slip events represent slip on a fault, for instance the plate boundary between a tectonic plate subducting under another tectonic plate. However, they take a much longer time (several days to several years) to happen relative to ordinary earthquakes. They have a relatively short recurrence time (months to years) compared to the recurrence time of regular earthquakes (up to several hundreds of years), allowing scientists to observe and study many complete event cycles, which is typically not possible to explore with traditional earthquake catalogs ~\citep{BER_2011}. A slow slip event on the plate boundary is inferred to happen when there is a reversal of the direction of motion at GNSS stations, compared to the secular interseismic motion. Slow slip events have been observed in many places ~\citep{BER_2011,AUD_2016}, such as Cascadia ~\citep{BAR_2020}, Nankai ~\citep{NIS_2013}, Alaska ~\citep{LI_2016}, Costa Rica ~\citep{JIA_2012}, Mexico ~\citep{RAD_2012}, and New Zealand ~\citep{WAL_2020}. \\

In many places, tectonic tremor is also observed in relation to slow slip, but it is more abundant in some places ~\citep{HAL_2018}. Tremor is a long (several seconds to many minutes), low amplitude seismic signal, with emergent onsets, and an absence of clear impulsive phases. Tectonic tremor have been explained as a swarm of small, low-frequency earthquakes (LFEs) ~\citep{SHE_2007_nature}, which are small magnitude earthquakes (M $\sim$ 1) for which frequency content (1-10 Hz) is lower than for ordinary earthquakes (up to 20 Hz). In subduction zones such as Nankai and Cascadia, tectonic tremor observations are spatially and temporally correlated with slow slip observations ~\citep{OBA_2002,ROG_2003}. Due to this correlation, these paired phenomena have been called Episodic Tremor and Slip (ETS). However, this is not always the case. For instance, in northern New Zealand, tremor is more challenging to detect, and seems to be located downdip of the slow slip on the plate boundary ~\citep{TOD_2016}. In Alaska, the tremor zone only partially overlaps the long-term slow slip zone and there does not appear to be any temporal correlation between tremor and slow slip occurrence ~\citep{WEC_2016}. \\

In Cascadia, there are robust signals in both GNSS and tremor ~\citep{HAW_2013}. This is also the case in Nankai ~\citep{HIR_2008}, where tiltmeters are used instead of GNSS. It is thus possible to use tremor as a proxy to observe slow slip events that are not directly observed in the GNSS data. For instance, ~\citet{AGU_2009} studied 23 ETS events in Cascadia with more than 50 hours of tectonic tremor. For all these events, they computed both the GPS-estimated moment release and the cumulative number of hours of tectonic tremor recorded. They observed a linear relationship between moment release and number of hours of tremor for ETS events of moment magnitude 6.3 to 6.8. Based on this linear relationship, it is possible to infer the existence of smaller slow slip events of magnitude 5-6 occurring simultaneously with smaller tremor bursts of duration 1 to 50 hours occurring in between the big ETS events, and for which there is no detectable signal in the GPS data. \\

@@

We focused our study on northwest Washington State. For the GNSS data, we used the GPS time series provided by the Pacific Northwest Geodetic Array, Central Washington University. These are network solutions in ITRF2008 with phase ambiguities resolved. Solutions are computed with JPL/NASA orbits and satellite clocks. North, East, and Vertical directions are available. However, as the direction of the secular plate motion is close to the East direction, we only used the East direction of the GPS time series for the data analysis, as it has the best signal-to-noise ratio. The wavelet method works best with data with zero mean, and no sharp discontinuities; so we use the cleaned dataset, that is GPS times series with linear trends, steps due to earthquakes or hardware upgrades, and annual and semi-annual sinusoids signals simultaneously estimated and removed following ~\citet{SZE_2004}. For the tremor data, we used the tremor catalog from the Pacific Northwest Seismic Network (PNSN) ~\citep{WEC_2010}. \\

@@

The Discrete Wavelet Transform (DWT) is an orthonormal transform that transforms a time series $X_t \left( t = 0, \cdots , N - 1 \right)$ into a vector of wavelet coefficients $W_i \left( i = 0 , \cdots , N - 1 \right)$. If we denote $J$ the level of the wavelet decomposition, and the number of observations is equal to $N = n * 2^J$, where $n$ is some integer higher or equal to 1, the vector of wavelet coefficients can be decomposed into $J$ wavelet vectors $W_j$ of lengths $\frac{N}{2}$, $\frac{N}{4}$, ... , $\frac{N}{2^J}$, and one scaling vector $V_J$ of length $\frac{N}{2^J}$. Each wavelet vector $W_j$ is associated with changes on time scale $\tau_j = dt 2^{j - 1}$, where $dt$ is the time step of the time series, and corresponds to the filtering of the original time series with a filter with nominal frequency interval $\lbrack \frac{1}{dt 2^{j + 1}} ; \frac{1}{dt 2^j} \rbrack$. The scaling vector $V_J$ is associated with averages in time scale $\lambda_J = dt 2^J$, and corresponds to the filtering of the original time series with a filter with nominal frequency interval $\lbrack 0 ; \frac{1}{dt 2^{j + 1}} \rbrack$. Wavelet vectors can be further decomposed into details and smooths, which are more easily interpretable. We define for $j = 1 , \cdots , J$ the $j$th wavelet detail $D_j$, which is a vector of length $N$, and is associated to time scale $\tau_j = dt 2^{j - 1}$. Similarly, we can define for $j = 1 , \cdots , J$ the $j$th wavelet smooth $S_j$, which is a vector of length $N$, and is associated to scales $\tau_{j + 1} = dt 2^{j + 1}$ and higher. The basic idea is to reapply to $W_j$ the wavelet filter that was used to construct $W_j$ from the initial time series $X$. Together, the details and the smooths define the multiresolution analysis (MRA) of $X$:

@@

To better evaluate the number of true and false detections, we convert the wavelet details into binary time series. If the absolute value of the wavelet detail is higher that a threshold, we replace the value by 1 (for positive values) or -1 (for negative values), otherwise we replace the value by 0. We do this on both the wavelet details of the GPS data and of the tremor data. Then we decide that if both the GPS and the tremor time series take the value 1 (or both take the value -1), we have a true detection (true positive, TP). If the GPS and the tremor time series have opposite signs, or if the absolute value of the GPS time series is 1 but the value of the tremor time series is 0, we have a false detection (false positive, FP). If both time series take the value 0, we do not have detection (true negative, TN). If the GPS time series take the value 0, but the absolute value of the tremor time series is 1, we miss a detection (false negative, FN). We then define the sensitivity (true positive rate) and the specificity (equal to 1 minus the false positive rate) as:

@@

We can then evaluate the quality of the detections obtained with our method by plotting a receiver operating characteristic curve (ROC curve). The ROC curve is widely use for binary classification problems in statistics and machine learning. We calculate an ROC value by varying the values of the threshold (here the two thresholds used to convert the GPS and the tremor time series into binary time series), computing the corresponding values of the true positive rate and the false positive rate (equal to 1 minus the specificity), and plotting the true positive rate as a function of the false positive rate. If the classification was made randomly, all the points would fall on the first diagonal. If the classifier was perfect, the corresponding point would fall on the top left corner of the graph with true positive rate equal to 1 and false positive rate equal to 0. The bigger the area under the curve, the better the classification method is. \\

As the slow slip events are better seen on levels 6, 7 and 8 of the wavelet decomposition, we first add the wavelet details corresponding to levels 6 to 8, and transform the resulting time series into a binary time series. We apply this transform to both the GPS and the tremor time series with varying thresholds. We then plot the ROC curve on Figure 9, each dot representing a different threshold. The corresponding sums of the wavelet details for the GPS data and the tremor data are shown on Figure 10. We can see that there is a trade-off between sensitivity and specificity as we vary the threshold. If we decrease the false positive rate, we also decrease the number of true events detected. If we increase the number of true events detected, we also increase the false positive rate. In Figure 10, we have chosen thresholds for the GPS time series and the tremor time series such that the specificity is higher than 0.75, and the sensitivity is the highest possible, that is we have chosen the thresholds corresponding to the dot that is farthest from the diagonal, which is random. \\

@@

\item Figure 2. Top: Data from GPS station PGC5 without missing values (black) and with missing values replaced by the sum of a straight line and a Gaussian noise component (red) for two locations of the missing values (left and right). The corresponding ten details and smooths of the wavelet composition are shown in increasing levels for the original data (black) and for the missing values replaced by linear interpolation plus Gaussian noise (red).

@@

\item Figure 9. Same as Figure 6 but for the sum of the 6th, 7th and 8th level details. The thresholds are 0.8 (for the GPS) and 0.01 (for the tremor).

\item Figure 10. ROC curve for the sum of the 6th, 7th, and 8th level details of the wavelet decomposition. Each black dot represents the true positive rate of event detections and the false positive rate of event detections for a given pair of thresholds (for the GPS and for the tremor). The red cross marks the true positive rate and the false positive rate obtained with the thresholds used to make Figure 9.

@@

\begin{figure}
\noindent\includegraphics[width=5cm, trim={0cm 0cm 0cm 0cm},clip]{figures/DS_10.eps}
\caption{Top: Data from GPS station PGC5 without missing values (black) and with missing values replaced by the sum of a straight line and a Gaussian noise component (red) for two locations of the missing values (left and right). The corresponding ten details and smooths of the wavelet composition are shown in increasing levels for the original data (black) and for the missing values replaced by linear interpolation plus Gaussian noise (red).}
\label{pngfiguresample}
\end{figure}

@@

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/ROC_6-7-8.eps}
\caption{ROC curve for the sum of the 6th, 7th, and 8th level details of the wavelet decomposition. Each black dot represents the true positive rate of event detections and the false positive rate of event detections for a given pair of thresholds (for the GPS and for the tremor). The red cross marks the true positive rate and the false positive rate obtained with the thresholds used to make Figure 9.}
\label{pngfiguresample}
\end{figure}

>>>>

Slow slip events are a new feature discovered in the last two decades in many subduction zones thanks to recordings of the displacement of Earth's surface by dense Global Navigation Satellite System (GNSS) networks \textbf{~\citep{VER_2010,SCH_2010,JIA_2012,WAL_2012}}. As with ordinary earthquakes, slow slip events represent slip on a fault, for instance the plate boundary between a tectonic plate subducting under another tectonic plate. However, they take a much longer time (several days to several years) to happen relative to ordinary earthquakes. They have a relatively short recurrence time (months to years) compared to the recurrence time of regular earthquakes (up to several hundreds of years), allowing scientists to observe and study many complete event cycles, which is typically not possible to explore with traditional earthquake catalogs ~\citep{BER_2011}. A slow slip event on the plate boundary is inferred to happen when there is a reversal of the direction of motion at GNSS stations, compared to the secular interseismic motion. Slow slip events have been observed in many places \textbf{~\citep{BER_2011,AUD_2016}, such as Cascadia ~\citep{BAR_2020}, Nankai ~\citep{NIS_2013}, Alaska ~\citep{LI_2016}, Costa Rica ~\citep{JIA_2012}, Mexico ~\citep{RAD_2012}, and New Zealand ~\citep{WAL_2020}}. \\

In many places, tectonic tremor is also observed in relation to slow slip, but it is more abundant in some places \textbf{~\citep{HAL_2018}}. Tremor is a long (several seconds to many minutes), low amplitude seismic signal, with emergent onsets, and an absence of clear impulsive phases. Tectonic tremor have been explained as a swarm of small, low-frequency earthquakes (LFEs) ~\citep{SHE_2007_nature}, which are small magnitude earthquakes (M $\sim$ 1) for which frequency content (1-10 Hz) is lower than for ordinary earthquakes (up to 20 Hz). In subduction zones such as Nankai and Cascadia, tectonic tremor observations are spatially and temporally correlated with slow slip observations ~\citep{OBA_2002,ROG_2003}. Due to this correlation, these paired phenomena have been called Episodic Tremor and Slip (ETS). However, this is not always the case. For instance, in northern New Zealand, tremor is more challenging to detect, and seems to be located downdip of the slow slip on the plate boundary ~\citep{TOD_2016}. In Alaska, the tremor zone only partially overlaps the long-term slow slip zone and there does not appear to be any temporal correlation between tremor and slow slip occurrence ~\citep{WEC_2016}. \\

In Cascadia, there are robust signals in both GNSS and tremor \textbf{~\citep{HAW_2013}}. This is also the case in Nankai \textbf{~\citep{HIR_2008}}, where tiltmeters are used instead of GNSS. It is thus possible to use tremor as a proxy to observe slow slip events that are not directly observed in the GNSS data. For instance, ~\citet{AGU_2009} studied 23 ETS events in Cascadia with more than 50 hours of tectonic tremor. For all these events, they computed both the GPS-estimated moment release and the cumulative number of hours of tectonic tremor recorded. They observed a linear relationship between moment release and number of hours of tremor for ETS events of moment magnitude 6.3 to 6.8. Based on this linear relationship, it is possible to infer the existence of smaller slow slip events of magnitude 5-6 occurring simultaneously with smaller tremor bursts of duration 1 to 50 hours occurring in between the big ETS events, and for which there is no detectable signal in the GPS data. \\

@@

We focused our study on northwest Washington State. For the GNSS data, we used the GPS time series provided by the Pacific Northwest Geodetic Array, Central Washington University. These are network solutions \textbf{in ITRF2014 with phase ambiguities resolved with wide-lane phase-biases and orbits and satellite clocks provided by the Jet Propulsion Laboratory/NASA}. North, East, and Vertical directions are available. However, as the direction of the secular plate motion is close to the East direction, we only used the East direction of the GPS time series for the data analysis, as it has the best signal-to-noise ratio. The wavelet method works best with data with zero mean, and no sharp discontinuities; so we use the cleaned dataset, that is GPS times series with linear trends, steps due to earthquakes or hardware upgrades, and annual and semi-annual sinusoids signals simultaneously estimated and removed following ~\citet{SZE_2004}. For the tremor data, we used the tremor catalog from the Pacific Northwest Seismic Network (PNSN) ~\citep{WEC_2010}. \\

@@

The Discrete Wavelet Transform (DWT) is an orthonormal transform that transforms a time series $X_t \left( t = 0, \cdots , N - 1 \right)$ into a vector of wavelet coefficients $W_i \left( i = 0 , \cdots , N - 1 \right)$. If we denote $J$ the level of the wavelet decomposition, and the number of observations is equal to $N = n * 2^J$, where $n$ \textbf{is some integer greater than or equal to 1}, the vector of wavelet coefficients can be decomposed into $J$ wavelet vectors $W_j$ of lengths $\frac{N}{2}$, $\frac{N}{4}$, ... , $\frac{N}{2^J}$, and one scaling vector $V_J$ of length $\frac{N}{2^J}$. Each wavelet vector $W_j$ is associated with changes on time scale $\tau_j = dt 2^{j - 1}$, where $dt$ is the time step of the time series, and corresponds to the filtering of the original time series with a filter with nominal frequency interval $\lbrack \frac{1}{dt 2^{j + 1}} ; \frac{1}{dt 2^j} \rbrack$. The scaling vector $V_J$ is associated with averages in time scale $\lambda_J = dt 2^J$, and corresponds to the filtering of the original time series with a filter with nominal frequency interval $\lbrack 0 ; \frac{1}{dt 2^{j + 1}} \rbrack$. Wavelet vectors can be further decomposed into details and smooths, which are more easily interpretable. We define for $j = 1 , \cdots , J$ the $j$th wavelet detail $D_j$, which is a vector of length $N$, and is associated to time scale $\tau_j = dt 2^{j - 1}$. Similarly, we can define for $j = 1 , \cdots , J$ the $j$th wavelet smooth $S_j$, which is a vector of length $N$, and is associated to scales $\tau_{j + 1} = dt 2^{j + 1}$ and higher. The basic idea is to reapply to $W_j$ the wavelet filter that was used to construct $W_j$ from the initial time series $X$. Together, the details and the smooths define the multiresolution analysis (MRA) of $X$:

@@

To better evaluate the number of true and false detections, we convert the wavelet details into \textbf{trinary} time series. If the absolute value of the wavelet detail is higher that a threshold, we replace the value by 1 (for positive values) or -1 (for negative values), otherwise we replace the value by 0. We do this on both the wavelet details of the GPS data and of the tremor data. Then we decide that if both the GPS and the tremor time series take the value 1 (or both take the value -1), we have a true detection (true positive, TP). If the GPS and the tremor time series have opposite signs, or if the absolute value of the GPS time series is 1 but the value of the tremor time series is 0, we have a false detection (false positive, FP). If both time series take the value 0, we do not have detection (true negative, TN). If the GPS time series take the value 0, but the absolute value of the tremor time series is 1, we miss a detection (false negative, FN). We then define the sensitivity (true positive rate) and the specificity (equal to 1 minus the false positive rate) as:

@@

We can then evaluate the quality of the detections obtained with our method by plotting a receiver operating characteristic curve (ROC curve). The ROC curve is widely use for binary classification problems in statistics and machine learning. We calculate an ROC value by varying the values of the threshold (here the two thresholds used to convert the GPS and the tremor time series into \textbf{trinary} time series), computing the corresponding values of the true positive rate and the false positive rate (equal to 1 minus the specificity), and plotting the true positive rate as a function of the false positive rate. If the classification was made randomly, all the points would fall on the first diagonal. If the classifier was perfect, the corresponding point would fall on the top left corner of the graph with true positive rate equal to 1 and false positive rate equal to 0. The bigger the area under the curve, the better the classification method is. \\

As the slow slip events are better seen on levels 6, 7 and 8 of the wavelet decomposition, we first add the wavelet details corresponding to levels 6 to 8, and transform the resulting time series into a \textbf{trinary} time series. We apply this transform to both the GPS and the tremor time series with varying thresholds. We then plot the ROC curve on Figure 9, each dot representing a different threshold. The corresponding sums of the wavelet details for the GPS data and the tremor data are shown on Figure 10. We can see that there is a trade-off between sensitivity and specificity as we vary the threshold. If we decrease the false positive rate, we also decrease the number of true events detected. If we increase the number of true events detected, we also increase the false positive rate. \textbf{If we increase the threshold for the tremor, the curve goes farther away from the first diagonal, that is we get better classification results. If we increase the threshold for the GPS, the false positive rate and the the number of events detected decrease.} In Figure 10, we have chosen thresholds for the GPS time series and the tremor time series such that the specificity is higher than 0.75 \textbf{(that is the false positive rate is lower than 0.25)}, and the sensitivity is the highest possible, that is we have chosen the thresholds corresponding to the dot that is farthest from the diagonal, which is random. \\

@@

\item Figure 2. Top: \textbf{Original data from GPS station PGC5 (black) and same data where displacement values have been artificially removed at two different times (2010.97 on the left and 2011.48 on the right) and replaced by the sum of a straight line and a Gaussian noise component (red)}. The corresponding ten details and smooths of the wavelet composition are shown in increasing levels for the original data (black) and for the \textbf{data with displacement values removed and replaced} by linear interpolation plus Gaussian noise (red).

@@

\item Figure 9. ROC curve for the sum of the 6th, 7th, and 8th level details of the wavelet decomposition. Each dot represents the true positive rate of event detections and the false positive rate of event detections for a given pair of thresholds (for the GPS and for the tremor). The \textbf{black} cross marks the true positive rate and the false positive rate obtained with the thresholds used to make Figure \textbf{10}. \textbf{The values of the threshold are color-coded. Reds correspond to the lowest value of the threshold for the tremor (0.001), while oranges, greens, blues, purples correspond to increasing values of the threshold for the tremor (up to 0.01). The brightest colors correspond to the highest values of the threshold for the GPS (1.5), while the darker and darker colors correspond to decreasing values of the threshold for the GPS (0.1).}

\item Figure 10. Same as Figure 6 but for the sum of the 6th, 7th and 8th level details. The thresholds are 0.8 (for the GPS) and 0.01 (for the tremor).

@@

\begin{figure}
\noindent\includegraphics[width=4.5cm, trim={0cm 0cm 0cm 0cm},clip]{figures/DS_10.eps}
\caption{Top: \textbf{Original data from GPS station PGC5 (black) and same data where displacement values have been artificially removed at two different times (2010.97 on the left and 2011.48 on the right) and replaced by the sum of a straight line and a Gaussian noise component (red)}. The corresponding ten details and smooths of the wavelet composition are shown in increasing levels for the original data (black) and for the \textbf{data with displacement values removed and replaced} by linear interpolation plus Gaussian noise (red).}
\label{pngfiguresample}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/ROC_6-7-8.eps}
\caption{ROC curve for the sum of the 6th, 7th, and 8th level details of the wavelet decomposition. Each dot represents the true positive rate of event detections and the false positive rate of event detections for a given pair of thresholds (for the GPS and for the tremor). The \textbf{black} cross marks the true positive rate and the false positive rate obtained with the thresholds used to make Figure \textbf{10}. \textbf{The values of the threshold are color-coded. Reds correspond to the lowest value of the threshold for the tremor (0.001), while oranges, greens, blues, purples correspond to increasing values of the threshold for the tremor (up to 0.01). The brightest colors correspond to the highest values of the threshold for the GPS (1.5), while the darker and darker colors correspond to decreasing values of the threshold for the GPS (0.1).}}
\label{pngfiguresample}
\end{figure}

@@
